---
title: "Introduction to GUSMap"
author: "Timothy P. Bilton"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Introduction to GUSMap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: "../inst/REFERENCES.bib"
biblio-style: "apalike"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set( collapse = TRUE, comment = "#>", eval=TRUE)
library(GUSMap)
```

In this tutorial, we will describe how to use GUSMap to perform linkage mapping and explain some of the functionality available. 

For this tutorial, we will use a simulated dataset of a full-sib family (F1 population) consisting of 100 offspring, 3 chromosomes and 1000 SNPs to illustrate GUSMap. The file location to the VCF file (and associated pedigree file) for the dataset can be extracted using the `simDS` function.

```{r simDS, eval = T}
vcffile <- simDS() # extract filename
vcffile            # filename stored in object vcffile
```

Some important up front comments regarding GUSMap:

1. Having high read depth on the parents (e.g., sequencing more than once) can be hugely helpful to correctly infer the segregation type of more SNPs. However, for the progeny, it is of more value (and cost-efficient) to have more individuals in the family at lower depth than sequence fewer at higher depth.
2. GUSMap is designed for low coverage sequencing data which means that there is no need to pre-filter the data in terms of read depth. Furthermore, GUSMap includes standard filtering steps in terms of minor allele frequency (MAF), missing data and a segregation test (including a few other filters) which means one only needs to input a raw Variant Call Format (VCF) file obtained after the SNP calling processes into GUSMap.

In addition, some terms that we will use throughout this tutorial are:

* Maternal informative (MI) SNPs: SNPs where the maternal parent is heterozygous and the paternal parent is homozygous.
* Paternal informative (PI) SNPs: SNPs where the paternal parent is heterozygous and the maternal marent is homozygous.
* Both informative (BI) SNPs: SNPs where both parents are heterozygous.

More details of this termology and the mathematics underpinning the methods in GUSMap can be found in @bilton2018genetics1.

## Loading Data

At present, GUSMap can read in data stored in a Variant Call Format (VCF) file, provided that there is some form of allelic depth information (e.g., the number of reads for the reference and alternate alleles). This is done using the `VCFtoRA` function which takes in the name of a VCF file and converts it into what we call a reference/alternate (RA) file.

```{r loadData, eval=T}
# convert VCF file to an RA file
rafile <- VCFtoRA(infilename = vcffile$vcf, direct = "./", makePed=TRUE)
```

The arguments of the `VCFtoRA` function are:

1. `infilename`: The name of the VCF file. 
2. `direct`: Specifies which directory to write the RA file to (relative to the current working directory).
3. `makePed`: Specifies whether or not to initalize a pedigree file. Pedigree files are inportant for creating a full-sib family object and will be discussed later. 

Currently, `VCFtoRA` requires one of the fields **AD**, **AO and RO**, or **DP4** to be present in the VCF file to extract allelic depth information. One other thing to note is that `VCFtoRA` returns the file location of the created RA file.
```{r rafile, eval=T}
rafile # file path of RA file
```

The RA file created is a tab-delimited file with columns, **CHROM** (the chromosome name taken from the "#CHROM" column in the VCF file), **POS** (the position of the SNP taken from the "POS" column in the VCF file), and **SAMPLES** which consists of the sampleIDs used in the VCF (columns 10 and above in the VCF file). For example, the first five columns and rows of the RA file for the simulated dataset are:

|CHROM	|POS	   |P1_1  |P2_1 |P1_2 |
|:------|:-------|:-----|:----|:----|
|1	    |866499	 |25,0  |0,2  |9,0  |
|1	    |2070050 |8,0   |4,9  |9,0  |
|1	    |2232468 |1,0	  |0,1  |2,0  |
|1	    |3459263 |12,13 |0,0  |0,1  |

Note: Indels and multiple alternative alleles are removed, and ./. is translated into 0,0.

An RA file can then be loaded into R using the `readRA` function.
```{r readRA, eval=T}
RAdata <- readRA(rafile = rafile, sampthres = 0.01, excsamp = NULL)
```

The arguments of the `readRA` function are:

1. `rafile`: Name of the RA file to be read in.
2. `sampthres`: Specifies the minimum sample depth of an individual before it is removed.
3. `excsamp`: Specifies IDs of any samples in the RA dataset to be discarded (for problematic samples to be removed). The input for `excsamp` must correspond to the sample IDs in the RA file.

The RA data is now stored as an `RA` object.
```{r class_RAdata, eval=T}
class(RAdata)
```

Summary information of the RA object can be displayed by printing the object.
```{r RAdata, eval=T}
RAdata
```

## Linkage mapping with GUSMap

In this section, we will discuss linkage mapping using GUSMap. The main steps to the process are:

1. Create a full-sib family and perform appropriate filtering
2. Constructing linkage groups
3. Ordering markers across each linakge group
4. Computing maps for the ordered linkage groups

In the following sections, we shall discuss how each of these steps are performed in GUSMap.

### Constructing a full-sib family

Construction of a 'full-sib' family (e.g., F1 population) object is achieved using the `makeFS` function.
```{r makeFS, eval=T}
mySpecies <- makeFS(RAobj = RAdata, pedfile = vcffile$ped, 
                      filter = list(MAF = 0.05, MISS = 0.5,
                        BIN = 100, DEPTH = 5, PVALUE = 0.01))
```

The `makeFS` function takes an `RA` object and pedigree file, creates a full-sib population and performs filtering on the data. 

For this tutorial, we use the existing pedigree file stored in the package. The filename of this pedigree file is stored in the object `vcffile$ped`. More information regarding the format of the pedigree file can be found by looking at the package documentation for the `makeFS` function which can be accessed by:
```{r help_makeFS, eval=F}
?`$makeFS`
```

There are five different types of filtering available in GUSMap:

* Minor allele frequency (`MAF`): SNPs are removed if the MAF is below the threshold since in full-sib families, the expected allele frequency are 0.5 for BI SNPs and 0.75 (or 0.25) for PI and MI SNPs.
* Percentage of missing data (`MISS`): SNPs are removed if the number of individuals without a single read for a given SNP is less than the threshold. 
* SNP binning distance (`BIN`): SNPs are binned together if the distance (in base pairs) between them is less than the threshold value. One SNP is then selected at random from each bin. This filtering is to ensure that there is only one SNP on each read left in the dataset.
* Parental depth (`DEPTH`): SNPs are removed if the read depth of each parent is below the threshold value. This filtering is performed since low read depth can result in parental heterozygous genotypes to be seen as homozygous resulting in incorrectly inferred segreation type.  
* Segregation test (`PVALUE`): A segregation test is performed to removed any SNPs in which the segregation type has been incorrectly inferred (e.g., parental genotypes called wrongly).

Again, more detials of these various filter types is available in the package documentation of the `makeFS` function.

### Linkage Groups


##### Computing 2-point recombination fraction estimates

Linkage group formation in GUSMap is based on 2-point recombination fraction estimates and associated LOD scores between all pairs of SNPs. These are computed using the function `$rf_2pt`. 
```{r 2pt_rf}
mySpecies$rf_2pt(nClust = 3)
```

The `$rf_2pt` function is parallelized to reduce computational time, where the `nClust` argument sepcifies how many cores to use in the parallelization. Be careful not to set this to more than what is available on your computer. Note: this step can take some time, especially if there are a large number of SNPs in your dataset. 

Once the 2-point recombination fraction (rf) estimates and associated LOD scores have been computed, we can plot the resulting rf estimates on a heatmap.
```{r plotChr}
# plot 2-point rf matrix for BI and MI SNPs
mySpecies$plotChr(mat = "rf", parent = "maternal", lmai=0.5)
```

Heatmaps are helpful for examining genome assemblies if the SNPs where called using a reference genome. For our example, we see that the middle section of chromosome 2 should actually be with chromosome 3, and that the last quarter of chromosome 1 needs to be inverted (that is the marker order need to be reversed). Furthermore, there are a number of SNPs that are clearly on the wrong chromosome and should be mapped to one of the other chromosomes. For *de novo* assemblies, one can still plot the 2-point rf's estimates although the resulting plot should look fairly random and so is usually not very informative.

One can also plot the matrix of 2-point LOD scores.
```{r plotChr2}
# plot 2-point LOD matrix for BI, PI and MI SNPs
mySpecies$plotChr(mat = "LOD", parent = "both", lmai=0.5)
```

In the above code, the `parent` argument has been set to `"both"` which means that all the SNPs will be plotted. The resulting plot looks a somewhat patchy. This is because the LOD scores between PI and MI SNPs are zero as there is no information regarding recombinaiton between these SNPs. The same also applies to the matrix of 2-point rf estimates.

##### Forming linkage groups

Using our 2-point rf estimates and LOD scores, we can proceed to forming linkage groups. 
```{r createLG}
mySpecies$createLG(parent = "both", LODthres = 5, nComp = 10)
mySpecies
```

The performance of the linkage grouping algorithm is controlled by the parameters `LODthres` and `nComp`. Details of these arguments and how the linkage groups are formed is explained in the package documentation at:
```{r help_createLG, eval=F}
?`$createLG`
```

Once the linkage groups have been formed, it is a good idea to look how well the grouping has performed by examining the heatmap of the 2-point rf estimates with the SNPs grouped according to the linkage groups. This can be done using the `$plotLG` function.
```{r plotLG_mat}
mySpecies$plotLG(parent = "both", interactive = F, LG = NULL)
```

The heatmap shows that there are four MI linkage groups and two PI linkage groups. We would expect there to be three MI linkage groups and three PI linkage groups since there are 3 chromosomes, and thus there is one extra MI linkage group and one missing PI linkage group. In general, the linkage between SNPs from different linkage groups appears relately low with high rf estimates. However,  we see that linkage group 2 has high rf with linkage group 6 which suggests that these two linkage groups should be joined together. Since linkage group 6 has many more SNPs that linkage group 2, it seems most likely that the segregation type for the SNPs in linkage group 2 were incorrectly inferred to be MI instead of PI (possibly due to sequencing errors and/or low read depth). Lastly, the heatmap suggests that linkage group 5 is actually two separate linkage groups (thus the missing linkage group) and so need to be split up. 

Note:

* Setting `interactive=TRUE` for the `$plotLG` function results in an interactive being plotted. 
Interactive plots make interrogating the rf (or LOD) plot and finding problematic SNPs easier. We have set this to `FALSE` 
in this tutorial to keep the size of this file small. In your session, try producing the interactive plot.
* At present, interactive plots generally work on a small number of SNPs (usually no more than 1000), so depending on 
the size of your linkage groups, you may need to plot each linkage group separately. You can use the `LG` argument to plot specific linkage groups.


##### Editing Linkage Groups

There are number of functions in GUSMap for editing linkage groups. Ideally, one should obtain as many linkage groups as the number of chromosomes the species is known to have. However, in practice, the number of linkage groups may be more, or less, than the number of chromosomes depending on the LOD threshold used and whether there is enough linkage between SNPs on the same chromosome to group all the SNPs located on the same chromosome together in the same linkage group. 

For our simulated data, we had four MI linkage groups (one more than the number of chromosomes being 3) and only two PI linkage groups. From the heatmaps above, we see that linkage group 5 appears to be two separate linkage groups and so needs to be split up. The best way to do this is recreating the paternal linkage groups using a higher LOD threshold.
```{r createLG_pat}
mySpecies$createLG(parent = "paternal", LODthres = 10, nComp = 10)
```

Note: the `parent` argument is set to `"paternal"` which means that we are only recreating the PI linkage groups. With a LOD threshold of 10, the number of PI linkage groups becomes four. Again, we examine the heatmap.
```{r plotLG_pat2}
mySpecies$plotLG(parent = "both", interactive = F)
```

So, the second PI linkage group (linkage group 6) with 2 SNPs appears to have some linkage with SNPs in linkage group 5. However, the linkage is only with a few SNPs and has rf values close to 0.5 with most of the other SNPs, so the linkage between linkage groups 5 and 6 may be spurious. There are two options that should be considered here; to merge linkage groups 5 and 6 or to discard linkage group 6. We will take the more caution option and remove linkage group 6. There are two ways of achieving this. One is to remove the entire linkage group.
```{r removeLG}
mySpecies$removeLG(LG = 6)
mySpecies
```

Alternatively, one could have also remove the SNPs in the linkage group.
```{r removeSNP, eval=FALSE}
mySpecies$removeSNP(snps = c(526,527))
```

Note that the `$removeSNPs` function is more useful for removing SNP(s) from a linkage group if they appear problematic whitout deleting the entire linkage group. This will be demonstrated later.

Figure ? also suggests that linkage group 2 should be combined with what was linkage group 6 but is now linkage group 7. We can merge these two linkage groups using the `$mergeLG` function.
```{r mergeLG}
mySpecies$mergeLG(LG = c(2,7), mergeTo="paternal")
mySpecies
```

Since linkage groups 2 and 7 are from different parental lines,, the `mergeTo` argument specifies whether the resulting merged linkage group should be considered as PI SNPs or MI SNPs.

**Note:** At this stage there is no undo funcationality in GUSMap. So, once you have removed a SNP or linkage group, or merge two linkage groups, you cannot reverse what you have done. However, in the case where you want to undo edits to your linkage group, then the best thing is to recreate the linkage groups using the `$createLG` function and proceed as you were doing before (and hence it helps to save you code in a script).

##### Adding Both-Informative SNPs

When the maternal and parental linkage groups have been tidied up, we can proceed to map the BI SNPs to the linkage groups and merge the linkage groups across the two parental lines. 
```{r addBIsnps}
mySpecies$addBIsnps(LODthres = 10, nComp = 10)
```

The resulting output shows that the PI and MI linkage groups have been merged together to form three complete linkage groups that include BI SNPs. Once again the linkage groups should be examine using heatmaps of the 2-point rf estimates and editing accordingly if required.

Note: The algorithm used in `$addBIsnps` is similar to the algorithm used for creating the paternal and maternal linkage groups (see the package documentation via `?$addBIsnps` for more details), and so the arguments `LODthres` and `nComp` have the same functionality as in the `$createLG` function. 

### Ordering Linkage Groups

Once we are happy with the linkage groups that we have formed, we can proceed to order the SNPs in each linkage group. In GUSMap, marker ordering is performed using the multidimensional scaling (MDS) approach as described by @preedy2016tag and is implmented in the `$orderLG` function.
```{r orderLG, , fig.width=8, fig.height=6}
mySpecies$orderLG(mapfun = "haldane", weight = "LOD2", ndim = 30) 
```

The two plots produced by the `mySpecies$orderLG` function are:

1. The first two dimensions of the components of the MDS.
2. The heatmap of the 2-point rf for all SNPs on the linkage group ordered according to the MDS algorithm.

These plots are helpful for examining how well the alogrithm has ordered the SNPs. Details of the MDS alogrithm are given in @preedy2016tag and information for the arguments `mapfun` and `weight` can be found in the package documentation (The default arguments tend to give the best results). 

```{r help_orderLG, eval=F}
?`$orderLG`
```

At this stage, it is again a good idea to examine the linkage groups using the `plotLG` function and discard any SNPs which appear to be problematic. In GUSMap, individual SNPs can be removed from the linakge groups using the `$removeSNP` function. From our analysis, we decided to remove the following SNPs.
```{r removeSNP_2}
mySpecies$removeSNP(c(
  c(253,255,300,316,143,301,340,166,173,201,209,264,312,310,315,336,356,64,500,459,513,449,423,376,451,508,418,228),
  c(377,405,217,177,436,450,467,28,515,411,229,222,410,230,401,368,375,34,259,164,169,334,283,213,148,288,213),
  c(112,108,311,37,69,136,46,115,29,126,42,72,26,92,67,71,94))
)
```

### Compute Linakge Maps

Now that we have finished tidying up our linkage groups, we proceed to compute the linkage maps using the `$computeMap` function.
```{r computeMap}
mySpecies$computeMap(chrom = NULL, nThreads = 2)
```

The `$computeMap` function has a number of arguments but the main ones are `chrom` which specifies which linkage group to compute 
linkage maps for and `nThreads` which specifies the number of threads to use in the computation of the likelihood (using OpenMP if available).
The resulting output gives the overall map distances and sequencing error estimates for each chromosome (in addition to the 
linkage group information). 

GUSMap have a couple of functions for quickly examining the linkage maps computed. The first function is to plot the linkage maps.
```{r plotLG}
mySpecies$plotLM(LG = NULL, fun = "haldane", col=c("cyan","lightblue","blue"))
```

The arguments for teh `$plotLM` function are:

* `LG`: The indices of the linkage groups to be plotted.
* `fun`: The mapping function to use. Currently only the Haldane, Kosambi and Morgan mapping functions are available.
* `col`: A vector of colors for the linkage maps.

In addition, we can also examine the order of the SNPs across the linkage groups relative to the original assembly.
```{r plotSyn, fig.width=7, fig.height=7}
mySpecies$plotSyn()
```

From the synteny plot, we see linkage group 1 corresponds to chromsome 3, linkage group 2 to chromosome 2 and linkage group 3 to chromosome 1. Furthermoe, the plot shows that the linkage mapping analysis has moved the middle section of the chromosome 2 to chromosome 1 and that it has inverted the SNP order at the end of chromosome 1.

Finally, one can extract the results of the linkage analysis in GUSMap by writing the resutls to a file using the `$writeLG` function.
```{r writeLG, eval=F}
mySpecies$writeLG(file = "mySpecies")
```

This is allows users to produce their own custom plots according in whatever software they are familiar with. For details of the information given in the file, see the package documentation.
```{r help_writeLG, eval=F}
?`$writeLG`
```


# References


